---
title: "Final project STAT5060"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2023-11-23"
editor_options:
  markdown:
    wrap: 72
---

# Data Introduction

The dataset "Nurses.csv" is from [the third edition of Multilevel
analysis: Techniques and applications (Hox, J. J., Moerbeek, M., &
Schoot, R,
2018)](https://github.com/MultiLevelAnalysis/Datasets-third-edition-Multilevel-book),
which contains data from a cross-sectional hypothetical study on stress
experienced by nurses in hospitals. In each of 25 hospitals, four wards
were selected and randomly assigned to an experimental or control
condition, where the intervention included a training program to cope
with job-related stress. After the intervention was implemented, a
random sample of 10 nurses from each ward completed a survey to test
job-related stress level.

Variables in the dataset:

*hospital*: ID number for a hospital $(1, \ldots, 25)$, a total of 25
hospitals

*ward*: ward numbering within a hospital $(1,2,3,4)$, a total of 4 wards
from each hospital

*wardid*: ID number of ward (not used in programming of analyses)

*nurse*: ID number for a nurse $(1, \ldots, 1000)$, a total of 1000
nurses (not used in programming)

*age*: nurse's age in years gender: nurse's gender $(0=$ male, $1=$
female $)$

*experience*: nurse's years of experience working stress: outcome
variable, a scale from 1 to 7

*wardtype*: type of hospital ward ('general care', 'special care')

*hospize*: hospital size ('small', 'medium', 'large')

*expcon*: intervention indicator variable ('control', 'experiment')
administered at Ward

# Basic data manipulations

```{r}
library(ggplot2)
library(nlme)
library(lme4)
library(mgcv)
library(readr)
library(furniture)    # nice table1() descriptives
library(stargazer)    # display nice tables: summary & regression
library(tidyverse)
library(dplyr)
nurses <- read_csv("./Nurses.csv")
nurses$hospsize.n<-recode(nurses$hospsize, 'small'=0, 'medium'=1, 'large'=2)
nurses$wardtype.n<-recode(nurses$wardtype, 'general care'=0, 'special care'=1)
nurses$expcon.n<-recode(nurses$expcon, 'control'=0, 'experiment'=1)
nurses$cgender<-ifelse(nurses$gender==0, 'Male', 'Female')
nurses$age.C <- nurses$age-mean(nurses$age)#mean of 43.0005 years
nurses$experience.C <- nurses$experience-mean(nurses$experience)#mean of 17.057 years
nurses$Chospsize <-(nurses$hospsize.n-1)
nurses$Cexpcon <- (nurses$expcon.n-0.5)
nurses = data.frame(nurses)
head(nurses)
```

# Exploratory Data Analysis
```{r}
stargazer(data.frame(nurses), title  = "Descriptive statistics, aggregate over entire sample",
                       header = FALSE,
                       type = "text")
```

```{r}
library(ggplot2)
p1 <- ggplot(nurses, aes(x=age, y=stress))
p1 + geom_smooth(aes(group=expcon, color=expcon), method="loess", se=FALSE) + 
  labs(title="Observed Stress versus Nurse's Age by Intervention Group",
       x="Nurse's Age (yrs)", y="Observed Stress")

```

-   The relationship between Nurse's age and Stress can be described as
    either linear if we assume that the smoothing was too smooth or as a
    bimodal (decreasing through age 40, increasing between age 40 and 52
    years, and decreasing after that). I will use a linear relationship
    between because there are fewer nurses in the older ages, and that
    could just be an artifact of the data.
-   Given that both lines are simply parallel to each other, there is no
    moderating effect of intervention on this association.

```{r}
p2 <- ggplot(data=nurses, aes(y=stress, x=cgender, fill=expcon))
p2 + geom_boxplot() + 
  labs(title="Distribution of Observed Stress by Gender and Intervention Group", 
       x="Gender", y="Observed Stress")

```

-   There may or may not be an association between gender and stress, as
    on average, stress seems to be very similar between intervention and
    control groups.
-   Because among males, there is a large overlaps in stress, and it is
    unclear whether there is an intervention effect among females, the
    visual inspection of the plot does not suggest a moderating effect
    of intervention on the association between stress and gender.

```{r}
p3 <- ggplot(nurses, aes(x=experience, y=stress))
p3 + geom_smooth(aes(group=expcon, color=expcon), method="loess", se=FALSE) + 
  labs(title="Observed Stress versus Nurse's Work Experience by Intervention Group",
       x="Nurse's Work Experience (yrs)", y="Observed Stress")

```

-   There appears to be a liner relationship between observed stress and
    nurse's experience in years.

-   Given that the two trajectories behave in parallel, this does not
    suggest a moderating effect of intervention on this association.

```{r}
p4 <- ggplot(data=nurses, aes(y=stress, x=wardtype, fill=expcon))
p4 + geom_boxplot() + 
  labs(title="Distribution of Observed Stress by Hospital Ward Type and Intervention Group", 
       x="Hospital Ward Type", y="Observed Stress")

```

-   Stress does not seem to depend on ward type, as if we were to
    average the control and intervention boxplots within each ward type,
    they would be approximately the same;

-   Since the intervention effect size seems to be the same within each
    ward type, therefore, there does not seem to be a moderating effect.

```{r}
p5 <- ggplot(data=nurses, aes(y=stress, x=hospsize, fill=expcon))
p5 + geom_boxplot() + 
  labs(title="Distribution of Observed Stress by Hospital Size and Intervention Group", 
       x="Hospital Size", y="Observed Stress")

```

-   If you average within each hospital size the effect size of
    intervention, it seems that there is a linear decline in stress
    across the hospital size;
-   Because the effect size of intervention is different between the
    hospital sizes, there is a potential moderating effect.

# Multi-level models
## 

## Three-level model using only a random intercept at each appropriate level of hierarchy

$$
\begin{aligned}
& \text { Level } 1 \text { (nurse): } Y_{i j k}=\gamma_{0 j k}+\gamma_{1 j k} \operatorname{Sex}_{i j k}+\gamma_{2 j k} \operatorname{Cexp}_{i j k}+\gamma_{3 j k} \operatorname{Cage}_{i j k}+\epsilon_{i j k} \\
& \text { Level } 2 \text { (ward): } \gamma_{0 j k}=\gamma_{00 k}+b_{0 j k}+\gamma_{01 k} \text {Nwardtype}_{j k}+\gamma_{02 k} \text {Cexpcon}_{i k} \\
& \text { Level } 3 \text { (hospital): } \gamma_{00 k}=\beta_{000}+b_{0 k}+\beta_{001} \text {Chospsize}_k \text {; } \\ \\
& \gamma_{02 k}=\beta_{020}+\beta_{021} \text {Chospsize}_k ; \gamma_{01 k}=\beta_{010} ; \gamma_{1 j k}=\beta_{100} ; \gamma_{2 j k}=\beta_{200} ; \gamma_{3 j k}=\beta_{300} \\ \\
& \text { Combined Model: } Y_{i j k}=\beta_{000}+b_{0 k}+b_{0 j k}+\beta_{001} \text {Chospsize}_k+ \beta_{010} \text {Nwardtype}_{i k} 
\\& +\left(\beta_{020}+\beta_{021} \text {Chospsize}_k\right) \operatorname{Cexpcon}_{j k}+\beta_{100} \text {Sex}_{i j k}+\beta_{200} \operatorname{Cexp}_{i j k}+\beta_{300} \operatorname{Cage}_{i j k}+\epsilon_{i j k} \\ 
&
\end{aligned}
$$

```{r}
m1 <- lme(stress ~   age.C + gender + experience.C + wardtype.n + Chospsize*Cexpcon, random = ~ 1+ward|hospital, 
               data=nurses, na.action=na.omit, method="ML")
summary(m1)
```

```{r}
VarCorr(m1)
```

```{r}
icc1 <- (as.numeric(VarCorr(m1)[2])+as.numeric(VarCorr(m1)[4]))/(as.numeric(VarCorr(m1) [2])+as.numeric(VarCorr(m1)[4])+as.numeric(VarCorr(m1)[5]) )
icc1
```

```{r}
icc2<-as.numeric(VarCorr(m1)[2])/(as.numeric(VarCorr(m1)[2])+as.numeric(VarCorr(m1)[4
])+ as.numeric(VarCorr(m1)[5]))
icc2
```

-   The estimated correlation among responses (stress level) within the
    same ward and the same hospital is 58.42%.

-   The estimated correlation among responses (stress level) from the
    same hospital regardless of ward type is 26.20%.

## Three-level model adding a random slope for centered intervention indicator variable due to hospital

$$
\begin{gathered}
\text { Level } 1 \text { (nurse): } Y_{i j k}=\gamma_{0 j k}+\gamma_{1 j k} \text {Sex}_{i j k}+\gamma_{2 j k} \text {Cexp}_{i j k}+\gamma_{3 j k} \text {Cage}_{i j k}+\epsilon_{i j k} \\
\text { Level } 2 \text { (ward): } \gamma_{0 j k}=\gamma_{00 k}+b_{0 j k}+\gamma_{01 k} \text {Nwardtype}_{i k}+\gamma_{02k} \text { Cexpcon }_{i k} \\
\text { Level } 3 \text { (hospital): } \gamma_{00 k}=\beta_{000}+b_{0 k}+\beta_{001} \text {Chospsize}_k ; \\\\
\begin{array}{c}
\gamma_{02 k}=\beta_{020}+\beta_{021} \text {Chospsize}_k+b_{02 k} ; \gamma_{01 k}=\beta_{010} ; \gamma_{1 j k}=\beta_{100} ; \gamma_{2 j k}=\beta_{200} ;
\gamma_{3 j k}=\beta_{300} \\\\
\text { Combined Model: } Y_{i j k}=\beta_{000}+b_{0 k}+b_{0 j k}+\beta_{001} \text {Chospsize}_k+ \beta_{010} \text {Nwardtype}_{j k}+ \\ d
\left(\beta_{020}+\beta_{021} \text {Chospsize}_k+b_{02 k}) \text {Cexpcon}_{j k}+\beta_{100} \text {Sex}_{i j k}+\beta_{200} \text {Cexp}_{i j k}\right. +\beta_{300} \text {Cage}_{i j k}+\epsilon_{i j k}
\end{array}
\end{gathered}
$$

```{r}
m2 <- lme(stress ~age.C + gender + experience.C + wardtype.n + Chospsize*Cexpcon, random = list(hospital=~1+Cexpcon, ward=~1), 
          data=nurses, na.action=na.omit, method="ML")
summary(m2)
```

```{r}
VarCorr(m2)
```

```{r}
icc2.1 = (0.166 + 0.203 + 0.111) / (0.166 + 0.203 + 0.111 + 0.217)
icc2.2 = (0.166 + 0.111) / (0.166 + 0.203 + 0.111 + 0.217)
icc2.3 = (0.166) / (0.166 + 0.203 + 0.111 + 0.217)
icc2.1
icc2.2
icc2.3
```

```{r}
icc2<-as.numeric(VarCorr(m2)[2])/(as.numeric(VarCorr(m2)[2])+as.numeric(VarCorr(m2)[4
])+ as.numeric(VarCorr(m2)[5]))
icc2
```
```{r}
intervals(m2)
```

-   Estimated variance of $b_{0k}$ is 0.1516 and it describes the variability
    in stress across different hospitals (Level 3 variance component);

-   Estimated variance of $b_{0jk}$ is 0.1087 and it describes the
    variability in stress across different wards within the same
    hospital (Level 2 variance component);

-   Estimated variance of $b_{02k}$ is 0.1784 describes variability in
    intervention effects across hospitals (Level 3 variance component);

-   Estimated correlation between $b_{0k}$ and $b_{02k}$ is 0.34, which
    means the larger the deviation of a hospital's mean for nurses'
    stress from the overall mean of stress averaged across all
    hospitals, the larger the deviation for that hospital's intervention
    effect from the overall intervention effect size averaged across all
    hospitals.

-   Estimated residual variance of 0.2166 is the estimated variance of
    epsilon and it captures the variability of the individual nurse's
    stress levels around the mean stress level for a specific ward
    within a hospital.

Similar to M1, the majority of heterogeneity in stress (response
variable) is explained by the correlation among responses from the same
ward within a hospital, but it is now composed of an additional source
of heterogeneity due to the varying effect sizes of intervention across
different hospitals: the variance of $b_{0k}$ increased from 0.1361 (M1)
to 0.1516 (M2), the variance of $b_{0jk}$ decreased from 0.1676 (M1) to
0.1087 (M2) but this decrease is compensated by the additional variance
of 0.1783 from the random effect of the heterogeneity of intervention
effect on stress (which is a ward-level covariate) across different
hospitals.

## Comparison between two models

To compare the two models, the only difference between them is in the
covariance matrix -- therefore, we need to conduct a LRT using REML:

The hypotheses are:

$$
\mathrm{H}_0: \sigma_{b_{02k}}^2=0
$$
$$
\mathrm{H}_1: \sigma_{b_{02k}}^2 \neq 0
$$

```{r}
m1 <- lme(stress ~   age.C + gender + experience.C + wardtype.n + Chospsize*Cexpcon, random = ~ 1|hospital/ward, 
          data=nurses, na.action=na.omit)
m2 <- lme(stress ~age.C + gender + experience.C + wardtype.n + Chospsize*Cexpcon, random = list(hospital=~Cexpcon, ward=~1), 
                data=nurses, na.action=na.omit)
anova(m1, m2)
```

This is the likelihood ratio test (LRT), implemented via REML, with
df=2, G2=8.406 and p=0.0149 (smaller than 0.10), indicating that the
addition of the random effect of intervention across different hospitals
is significantly improving the fit of the model to the data.

Hence, we decide to choose model2.

```{r}
summary(m2)
```

# Interpretation of the final model

-   *Intercept*: estimated mean stress for a male nurse working on a
    general ward of a medium sized hospital, with average age and
    average years of experience, is 5.40 points;
-   *Age.c*: **for every one year of increase in age, mean stress level
    significantly increases by 0.02 points;**
-   *Gender*: **female nurses reported on average significantly lower
    stress than male nurses, and that difference is estimated to be 0.45
    points;**
-   *Experience.C*: **for every one year of increase in work experience,
    mean stress level significantly decreases by 0.06 points;**
-   *Wardtype.n*: while the estimated mean stress in a specialty care
    ward was on average higher than in the general ward by 0.05 points,
    **this difference was neither clinically nor statistically
    significant or meaningful;**
-   *Chospsize, Cexpcon and Chospsize*$\times$Cexpcon should be
    interpreted together: as hospital size increased from small to
    medium to large, the significant reduction in mean stress level due
    to intervention was significantly diminished/diluted: there was **a
    significant reduction in mean stress level due to intervention at
    small hospitals** (an estimated reduction of
    Beta6-Beta7=-0.499-0.998=-1.5), **a significant reduction in mean
    stress level due to intervention in medium-sized hospitals** (an
    estimated reduction of Beta6=-0.499 points), and a non-significant
    increase in large-sized hospitals (an estimated increase of
    Beta6+Beta7=-0.499+.998=0.499), meaning that **intervention did not
    seem to significantly impact mean stress level among nurses in large
    hospitals.**

# Residual and random effects visual diagnostic check

```{r}
m.finalb = m2
library(mgcv)
est.cov.mfinalb<-extract.lme.cov(m.finalb)
pred.m.finalb = fitted(m.finalb, level=0:2)
pred.m.finalb0 = fitted(m.finalb, level=0)
pred.m.finalb1=fitted(m.finalb, level=1)
pred.m.finalb2=fitted(m.finalb, level=2)
resid.m.finalb=resid(m.finalb, level=0:2)

cr.final <- solve(t(chol(est.cov.mfinalb))) %*% residuals(m.finalb, level=0)

pred.mfinalb = data.frame(id=nurses$nurse,ctrt=nurses$expcon,trt=nurses$expcon.n,   age=nurses$age,experience=nurses$experience, hospsize=nurses$hospsize.n,chospsize=nurses$hospsize, gender=nurses$gender, cgender=nurses$cgender, pred=fitted(m.finalb, level=0), type=nurses$wardtype.n, ctype =nurses$wardtype, resid=cr.final[,1])
```

## Histogram and a Q-Qplot for the transformed residuals

```{r}
library(gridExtra)
#Histogram of transformed residuals for final model
p1 <- ggplot(pred.mfinalb, aes(x=resid))
p1<- p1 + geom_histogram(aes(y=..density..), color="black", fill="lightblue") +
  geom_density() + labs(title="Distribution of Transformed Residuals-LME Final Model", x="Residual")

#Q-Q plot of transformed residuals for m3
p2 <- ggplot(pred.mfinalb, aes(sample=resid))
p2<- p2 + geom_qq_line(color="blue",size=1) + geom_qq() +
  labs(title="Q-Q plot of Transformed Residuals-LME Final Model", x="Quantile", y="Residual")

#install.packages("gridExtra")
#library("gridExtra")
plots.Q3di<-grid.arrange(p1,p2, nrow=1)
plots.Q3di

```

The histogram of residuals looks pretty normal, and the Q-Q plot
indicates no significant departures from the Normality assumption.

## Transformed residuals against the marginal mean response and each covariate in the final model

```{r}
p3 <- ggplot(pred.mfinalb, aes(x=pred, y=resid))
p3 <- p3 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Predicted Mean", y="Residual")

#Scatter plot of transformed residual against age with LOESS smoothed curve for final model
p4 <- ggplot(pred.mfinalb, aes(x=age, y=resid))
p4 <- p4 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Age (yrs)", y="Residual")

#Scatter plot of transformed residual against Experience with LOESS smoothed curve for final model
p5 <- ggplot(pred.mfinalb, aes(x=experience, y=resid))
p5 <- p5 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Experience (yrs)", y="Residual")

#Scatter plot of transformed residual against Hospital Size with LOESS smoothed curve for final model
p6 <- ggplot(pred.mfinalb, aes(x=hospsize, y=resid))
p6 <-p6 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Hospital Size, 0=Small, 1=Medium, 2=Large", y="Residual")

#Scatter plot of transformed residual against Gender with LOESS smoothed curve for final model
p7 <- ggplot(pred.mfinalb, aes(x=gender, y=resid))
p7 <- p7 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Gender, 0=Male, 1=Female", y="Residual")

#Scatter plot of transformed residual against Treatment with LOESS smoothed curve for final model
p8 <- ggplot(pred.mfinalb, aes(x=trt, y=resid))
p8 <- p8 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Treatment", y="Residual")

#Scatter plot of transformed residual against Ward Type with LOESS smoothed curve for final model
p9 <- ggplot(pred.mfinalb, aes(x=type, y=resid))
p9 <- p9 + geom_point() + geom_hline(yintercept=0,colour="red") + 
  geom_smooth(method="loess", se=F) +
  labs(x="Ward Type, 0=General Care, 1=Special Care", y="Residual")
plots.Q3dii<-grid.arrange(p3,p4,p5,p6,p7,p8,p9, nrow=3, top="Transformed Residuals by Covariates")
plots.Q3dii
plot(plots.Q3dii)
```

The distribution of transformed residuals relative to each covariate
indicates that the model for the mean response fits the data well and
that the functional form for each covariate seems to be appropriate, as
there are no systematic trends identified visually: the residuals are
approximately 0 on average and their variability is approximately half
above and half below the average of 0.

## Histograms and Q-Q plots of the random effects (BLUPS)

```{r}
###Assessing normality of random effects for final model
mfinalb.re <- random.effects(m.finalb)
str(mfinalb.re)

hosp.level.re<-as.data.frame(mfinalb.re$hospital)
head(hosp.level.re)
length(hosp.level.re$Cexpcon)#25 hospitals
ward.level.re<-as.data.frame(mfinalb.re$ward)
head(ward.level.re)
length(ward.level.re$`(Intercept)`)#100 wards

#Calculate the variance of the empirical BLUPs and compare them to the estimated variances for final model
apply(hosp.level.re, 2, var)
#(Intercept)     Cexpcon 
#0.130260    0.110292 
apply(ward.level.re, 2, var)
#(Intercept) 
#0.06012455 

#Histogram of transformed Random Effects for final model
p1 <- ggplot(hosp.level.re, aes(x=`(Intercept)`))
p1 <- p1 + geom_histogram(aes(y=..density..), color="black", fill="lightblue") +
  geom_density() + labs(title="Random Int for Hospital", x="Hospital RE")

#Q-Q plot 
p2 <- ggplot(hosp.level.re, aes(sample=`(Intercept)`))
p2 <- p2 + geom_qq_line(color="blue",size=1) + geom_qq() +
  labs(title="Random Int for Hospital", x="Quantile", y="Hospital RE")

#Histogram
p3 <- ggplot(hosp.level.re, aes(x=Cexpcon))
p3 <- p3 + geom_histogram(aes(y=..density..), color="black", fill="lightblue") +
  geom_density() + labs(title="Random Trt across Hospitals", x="Hospital*Treat RE")

#Q-Q plot 
p4 <- ggplot(hosp.level.re, aes(sample=Cexpcon))
p4 <- p4 + geom_qq_line(color="blue",size=1) + geom_qq() +
  labs(title="Random Trt across Hospitals", x="Quantile", y="Hospital*Treat RE")

#Histogram 
p5 <- ggplot(ward.level.re, aes(x=`(Intercept)`))
p5 <- p5 + geom_histogram(aes(y=..density..), color="black", fill="lightblue") +
  geom_density() + labs(title="Random Int for Ward", x="Ward RE")

#Q-Q plot 
p6 <- ggplot(ward.level.re, aes(sample=`(Intercept)`))
p6 <- p6 + geom_qq_line(color="blue",size=1) + geom_qq() +
  labs(title="Q-Q plot of Random Int for Ward", x="Quantile", y="Ward RE")

plots.Q3diii<-grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2, top="Distribution of Random Effects: Histogram and Q-Q Plots")
plots.Q3diii
plot(plots.Q3diii)
```

-   Random Intercept for Hospital: there do not appear to be fatal
    departures from normality for the random intercepts for Hospital,
    and their empirical distribution simply reflects the relatively
    small number of hospitals in the study (only 25 hospitals).

-   Random Treatment Effect across Hospitals: similarly to the
    distribution of RE for Hospital, this RE suffers from the small
    sample size of hospitals, and does not appear to have fatal
    departures from normality (Q-Q plot).

-   Random Intercept for Ward: there do not appear to be fatal
    departures from normality for the RE for ward; because there were
    more wards than hospitals, the histogram and the Q-Q plot for this
    RE looks better for the other two Res, but it's only due to the
    increased sample size (number of wards is 100).


